---
title: "Group 3 Final Report"
output: html_document
---


```{r, warning=FALSE, message=FALSE}
library(tidyverse)
library(rmarkdown)
library(lubridate)
library(ggthemes)
```

<!-- Start of Anders --->
# Overview of U.S. Motor Accidents

Before we dive into the specific features of crashes and possible mitigations,
we first take a look at the 4.2 million record accident dataset as a whole.

The dataset spans from 2016 to 2020. We first process the data into a form
that allows us to summarize some aspects of it more easily. 

We also use a chart to convert the state abbreviations to state names as well
as a dataset contains census population counts.

```{r, warning=FALSE, message=FALSE}
#acc_raw <- read_csv('US_Accidents_Dec20.csv')

# TEMP
acc_raw <- read_csv('US_Accidents_Sample.csv')

census_raw <- read.csv('census2010-2019.csv')
states <- map_data('state')
abv <- read.csv('state_abbrvs.csv', fileEncoding="UTF-8-BOM")


# === Census Preprocessing ===
# Separate area into city and state, summarize population per state in each year
census <- census_raw %>%
  mutate(across(Census:X2019, function(x){as.numeric(gsub(",", "", x))})) %>%
  separate(Geographic.Area, sep=', ', into=c('City', 'region')) %>%
  group_by(region) %>%
  summarize(
    `2016`=sum(X2016),
    `2017`=sum(X2017),
    `2018`=sum(X2018),
    `2019`=sum(X2019)
  ) %>%
  pivot_longer(`2016`:`2019`, names_to='Year', values_to='Population') %>%
  mutate(Year=as.numeric(Year), State=region) %>%
  select(-region)


# === Accidents Preprocssing ===
# Convert state code to state name, format time, select only the columns needed
acc <- acc_raw %>%
  mutate(Code=State) %>%
  select(-State) %>%
  left_join(abv, by='Code') %>%
  select(Start_Time, Start_Lat, Start_Lng, State) %>%
  mutate(State=tolower(State), Time=ymd_hms(Start_Time)) %>%
  select(-Start_Time) %>%
  mutate(State=str_to_title(State))

states <- states %>%
  mutate(State=str_to_title(region)) %>%
  select(-region, -subregion)
```

First, we can visualize the accidents all over the U.S. by plotting them as points over a map of the U.S.

```{r}
states %>% ggplot(aes(x=long, y=lat)) + 
  geom_polygon(aes(group=group)) +
  geom_point(data=acc, aes(x=Start_Lng, y=Start_Lat), color='magenta2', size=0.02, alpha=0.02) +
  coord_map()
```

We can see that accidents are concentrated in cities and also on roads, since of course
those are locations where cars drive the most, and therefore where accidents are most likely to occur.

However, we may want to instead see the total accidents by state. We can do so using a choropleth map.

Below is a choropleth map showing the total accidents over 2016-2020 by state.

```{r}
acc_states <- acc %>% group_by(State) %>% summarize(total_accidents=n())

acc_map <- acc_states %>% left_join(states, by='State')

ggplot(acc_map, aes(x=long, y=lat)) + 
  geom_polygon(aes(group=group, fill=total_accidents)) + 
  coord_map() + labs(fill='Total Accidents') + 
  scale_fill_gradient(low="#222222", high="magenta2") + 
  ggtitle('Total Accidents from 2016 to 2020 by State')
```

Above we can see that clearly the states with the highest populations (California, Texas, Florida) 
have the most accidents total, which makes sense. It might be more insightful to instead
look at the per capita accident rates.

```{r}
acc2019 <- acc %>% filter(year(Time)==2019)

census2019 <- census %>% filter(Year==2019)

acc2019 <- acc2019 %>% left_join(census2019, by='State')

acc2019_states <- acc2019 %>% 
  group_by(State) %>%
  summarize(total_accidents=n(), population=first(Population), per1000=n()*1000/first(Population))

acc2019_states %>% paged_table

acc_map_capita <- acc2019_states %>% left_join(states, by=c('State'))

ggplot(acc_map_capita, aes(x=long, y=lat)) + 
  geom_polygon(aes(group=group, fill=per1000)) + 
  coord_map() + ggtitle('Per Capita Accidents in 2019') + 
  labs(fill='Accidents per 1000 people') + 
  scale_fill_gradient(low="#222222", high="magenta2")
```

Interestingly, we can see that South Carolina and Oregon seem to have the highest per capita accident rates.

We can further investigate this with the use of a bar chart.

```{r}
col <- c(rep(c('grey50'), 8), c('red'), rep(c('grey50'), 40))

acc2019_states %>% 
  mutate(State=str_to_title(State)) %>%
  arrange(per1000) %>%
  mutate(State=factor(State, levels=State)) %>%
  ggplot(aes(x=State, weight=per1000, fill=State)) + geom_bar() + coord_flip() +
  scale_fill_manual(values=col) +
  ylab('Accidents per 1000 People in 2019') +
  theme(legend.position = "none")
```

We can see that South Carolina has approximately double the accident rate of Oregon.

It's difficult to attribute this to some cause. It could be some problem with South Carolina's 
infrastructure or driving education but that lies outside the scope of this project.

We can see Iowa highlighted in red. Iowa is clearly not the worst in terms of accident rate, as 
many states have far worse rates, but there is also still room for improvement as there
are a few states with lower accident rates than Iowa.
<!-- End of Anders --->
  
  
  
<!-- Start of Jack --->

# Accidents in Iowa

### Load Data Frames
```{r, message=FALSE}
census <- read_csv("census.csv")%>%
  filter(STNAME == "Iowa", PRIMGEO_FLAG == 0, FUNCSTAT == "A")
iowa <- read_csv("US_Accidents_Dec20.csv", col_types = cols(End_Lat=col_character(),End_Lng=col_character()))%>%
  select(State, Severity, Start_Time, Start_Lat, Start_Lng, `Wind_Speed(mph)`,Weather_Condition)%>%
  filter(State %in% "IA")

county <- map_data("county")
```



### Plot Accidents in Iowa

We also decided to look at accidents close to home. Evident in the map below, accidents in Iowa cluster in areas we know to be the most congested; namely, many accidents occur around Des Moines and along I-35 and I-80. Additionally, Iowa's northwestern corner appears to exhibit disproportionately many severe accidents relative to non-severe ones.

```{r}
iowacounties<- county%>%
  filter(region %in% "iowa")

p<-ggplot(iowacounties, aes(x=long, y=lat))+geom_polygon(aes(group=group))
p<-p+geom_point(data=iowa, aes(x=Start_Lng, y=Start_Lat, color=factor(Severity), alpha=0.15))
p<-p+labs(title="Accidents in Iowa", subtitle="Data from 2016 to 2020")+theme_map()
p
```


### Plot Severity by Wind Speed and Weather Type

Nearly any Iowan can testify to the hazards of slippery weather. However, we sought to quantify weather's influence on accident frequency and severity by observing crashes recorded in Iowa over the past five years.

```{r}
iowa1<-iowa%>%
  select(`Wind_Speed(mph)`,Severity, `Weather_Condition`)%>%
  na.omit()%>%
  group_by(Weather_Condition)%>%
  summarize(`Average Severity` = mean(Severity), `Average Wind Speed (mph)` = mean(`Wind_Speed(mph)`), n=n())%>%
  filter(n>15)
```

Slippery conditions are generally associated with the most severe crashes. Accidents in 'light freezing rain' were the most severe by a considerable margin, followed by snow, rain, and light thunderstorms. 

Based on the results, Iowans should limit driving on days with lots of precipitation and reduce their speeds. 

```{r}
ggplot(iowa1, aes(y=`Average Severity`, x=`Average Wind Speed (mph)`))+geom_text(aes(label=`Weather_Condition`), size=2)+labs(title="Severity and Wind Speed by Weather Type")
```



### Accidents in Iowa by Day of the Week
```{r}
iowa %>%
  mutate(Day = wday(Start_Time, label=TRUE, abbr=FALSE))%>%
  group_by(Day)%>%
  ggplot(aes(x=Day))+geom_bar()+labs(title="Accidents in Iowa by Weekday")
```

Iowans typically crash the most on weekdays. This result is pretty intuitive--people tend to drive more during the week.

<!-- End of Jack --->

<!-- Start of Eric --->

```{r}
acc = read.csv("US_Accidents_Dec20.csv")
```

```{r}
acc = separate(acc, Start_Time, into = c("Date", "Time"), sep = " ")
```

```{r}
acc$Date = ymd(acc$Date)
```

```{r}
acc$Year = year(acc$Date)
acc$Month = month(acc$Date)
acc$Day = day(acc$Date)
```

```{r}
daily_acc = acc %>% group_by(Date) %>% summarise(n = n()) %>% mutate(year = year(Date))
```

Here we had to separate the Start_Time column into time and date, then convert the date into Date format using the lubridate library. 

## 2016-2020 Daily Accidents in America

```{r}
daily_acc %>% ggplot(aes(x=Date, y=n)) + geom_line() + facet_wrap(~year, scales = "free_x") + scale_x_date(date_labels = "%b")
```

This is a time series of the number of daily accidents. We can see from this graph that the number of accidents remains fairly consistent over time, with only slight jumps, probably caused by adding more ways of obtaining accident data. 2020 is the year that is much different from the rest, with the number of accidents dipping during the pandemic, and then rising higher than the pre-pandemic numbers.

## 2020 Daily Accidents in America

```{r}
daily_acc %>% filter(year == 2020) %>% ggplot(aes(x=Date, y=n)) + geom_line() + facet_wrap(~year, scales = "free_x")
```

Here we can see a better look of what happened in 2020. This graph, along with some research, tells us that driving speeds increased during the pandemic, which led to more accidents once more cars returned to the road.

## Total Accidents by Month

```{r}
daily_acc %>% filter(year != 2016) %>% group_by(month(Date)) %>% summarise(n=sum(n)) %>% rename("month" = "month(Date)") %>% ggplot(aes(x=month, y=n)) + geom_bar(stat = "identity")
```

This graph shows us that July is the month with the fewest accidents from 2017 to 2020, and December is the month with the most. This is all due to the numbers in 2020, because the months were pretty even until then.

## 2016-2020 Severe Accidents in America

```{r}
acc %>% filter(Severity==4) %>% group_by(Date) %>% summarise(n = n()) %>% mutate(year = year(Date)) %>% ggplot(aes(x=Date, y=n)) + geom_line() + facet_wrap(~year, scales = "free_x") + scale_x_date(date_labels = "%b")
```

This is a look at the severe accidents (accidents with the highest severity score) per day. Here we can see that the numbers remained the same for the most part until the pandemic hit, when numbers then spiked.

## 2020 Severe Accidents in America

```{r}
acc %>% filter(Severity==4, Year==2020) %>% group_by(Date) %>% summarise(n = n()) %>% mutate(year = year(Date)) %>% ggplot(aes(x=Date, y=n)) + geom_line() + facet_wrap(~year, scales = "free_x") + scale_x_date(date_labels = "%b")
```

Here is a closer look at the year 2020, and we can see the large spike at the start of the pandemic. Research found that driving speeds increased with fewer cars on the road, so those higher speeds might have led to more severe accidents.

## 2016-2020 Accidents in Iowa

```{r}
acc %>% filter(State=="IA") %>% group_by(Date) %>% summarise(n = n()) %>% mutate(year = year(Date)) %>% ggplot(aes(x=Date, y=n)) + geom_line() + facet_wrap(~year, scales = "free_x") + scale_x_date(date_labels = "%b")
```

This chart is a look at the number of daily accidents in the state of Iowa. We can see from this graph that the drivers in Iowa performed much like the rest of the United States. There was a large spike in the number of accidents at the end of 2020.

## 2020 Accidents in Iowa

```{r}
acc %>% filter(Year == 2020, State=="IA") %>% group_by(Date) %>% summarise(n = n()) %>% mutate(year = year(Date)) %>% ggplot(aes(x=Date, y=n)) + geom_line() + facet_wrap(~year, scales = "free_x") + scale_x_date(date_labels = "%b")
```

Looking closer at 2020, we can see that spike at the end of 2020. This, along with the prior graph, tells us that the driving behavior of Iowans matches the driving behavior of the rest of America.


<!-- End of Eric --->

Conclusion:
	
Although motor vehicle crashes are calamitous and will remain a problem into the future, our report reveals key signs of hope. Namely, we identified disparities in the number and severity of crashes across locations, suggesting states with very high crash rates may be able to reduce their crash rate through policy intervention.

Our data set supported the finding that the COVID-19 significantly increased the number and severity of motor vehicle crashes. We noted that the severity of crashes initially increased, then lulled, before increasing quite steeply at the end of 2020. This suggests that drivers drove faster and more recklessly after the start of the pandemic and that they did not adjust their behavior as cars returned to the roads.

Contribution Statement:

Anders:

Eric: My analysis focused on the changes in number of daily accidents over the years. This analysis focused heavily on the impact of COVID-19, because the pandemic had a large impact on traffic, and therefore the number of accidents. I also contributed to the team report and presentation.

Jack: My analyses focused primarily on Iowa. I created a map of Iowa and two additional graphics to represent accidents by weekday and by weather type. I also contributed to this report as well as the project presentation.

Murat:

Ryan:


