---
title: "Group 3 Final Report"
output: html_document
---


```{r, warning=FALSE, message=FALSE}
library(tidyverse)
library(rmarkdown)
library(lubridate)
library(ggthemes)
```

<!-- Start of Anders --->
# Overview of U.S. Motor Accidents

Before we dive into the specific features of crashes and possible mitigations,
we first take a look at the 4.2 million record accident dataset as a whole.

The dataset spans from 2016 to 2020. We first process the data into a form
that allows us to summarize some aspects of it more easily. 

We also use a chart to convert the state abbreviations to state names as well
as a dataset contains census population counts.

```{r, warning=FALSE, message=FALSE}
#acc_raw <- read_csv('US_Accidents_Dec20.csv')

# TEMP
acc_raw <- read_csv('US_Accidents_Sample.csv')

census_raw <- read.csv('census2010-2019.csv')
states <- map_data('state')
abv <- read.csv('state_abbrvs.csv', fileEncoding="UTF-8-BOM")


# === Census Preprocessing ===
# Separate area into city and state, summarize population per state in each year
census <- census_raw %>%
  mutate(across(Census:X2019, function(x){as.numeric(gsub(",", "", x))})) %>%
  separate(Geographic.Area, sep=', ', into=c('City', 'region')) %>%
  group_by(region) %>%
  summarize(
    `2016`=sum(X2016),
    `2017`=sum(X2017),
    `2018`=sum(X2018),
    `2019`=sum(X2019)
  ) %>%
  pivot_longer(`2016`:`2019`, names_to='Year', values_to='Population') %>%
  mutate(Year=as.numeric(Year), State=region) %>%
  select(-region)


# === Accidents Preprocssing ===
# Convert state code to state name, format time, select only the columns needed
acc <- acc_raw %>%
  mutate(Code=State) %>%
  select(-State) %>%
  left_join(abv, by='Code') %>%
  select(Start_Time, Start_Lat, Start_Lng, State) %>%
  mutate(State=tolower(State), Time=ymd_hms(Start_Time)) %>%
  select(-Start_Time) %>%
  mutate(State=str_to_title(State))

states <- states %>%
  mutate(State=str_to_title(region)) %>%
  select(-region, -subregion)
```

First, we can visualize the accidents all over the U.S. by plotting them as points over a map of the U.S.

```{r}
states %>% ggplot(aes(x=long, y=lat)) + 
  geom_polygon(aes(group=group)) +
  geom_point(data=acc, aes(x=Start_Lng, y=Start_Lat), color='magenta2', size=0.02, alpha=0.02) +
  coord_map()
```

We can see that accidents are concentrated in cities and also on roads, since of course
those are locations where cars drive the most, and therefore where accidents are most likely to occur.

However, we may want to instead see the total accidents by state. We can do so using a choropleth map.

Below is a choropleth map showing the total accidents over 2016-2020 by state.

```{r}
acc_states <- acc %>% group_by(State) %>% summarize(total_accidents=n())

acc_map <- acc_states %>% left_join(states, by='State')

ggplot(acc_map, aes(x=long, y=lat)) + 
  geom_polygon(aes(group=group, fill=total_accidents)) + 
  coord_map() + labs(fill='Total Accidents') + 
  scale_fill_gradient(low="#222222", high="magenta2") + 
  ggtitle('Total Accidents from 2016 to 2020 by State')
```

Above we can see that clearly the states with the highest populations (California, Texas, Florida) 
have the most accidents total, which makes sense. It might be more insightful to instead
look at the per capita accident rates.

```{r}
acc2019 <- acc %>% filter(year(Time)==2019)

census2019 <- census %>% filter(Year==2019)

acc2019 <- acc2019 %>% left_join(census2019, by='State')

acc2019_states <- acc2019 %>% 
  group_by(State) %>%
  summarize(total_accidents=n(), population=first(Population), per1000=n()*1000/first(Population))

acc2019_states %>% paged_table

acc_map_capita <- acc2019_states %>% left_join(states, by=c('State'))

ggplot(acc_map_capita, aes(x=long, y=lat)) + 
  geom_polygon(aes(group=group, fill=per1000)) + 
  coord_map() + ggtitle('Per Capita Accidents in 2019') + 
  labs(fill='Accidents per 1000 people') + 
  scale_fill_gradient(low="#222222", high="magenta2")
```

Interestingly, we can see that South Carolina and Oregon seem to have the highest per capita accident rates.

We can further investigate this with the use of a bar chart.

```{r}
col <- c(rep(c('grey50'), 8), c('red'), rep(c('grey50'), 40))

acc2019_states %>% 
  mutate(State=str_to_title(State)) %>%
  arrange(per1000) %>%
  mutate(State=factor(State, levels=State)) %>%
  ggplot(aes(x=State, weight=per1000, fill=State)) + geom_bar() + coord_flip() +
  scale_fill_manual(values=col) +
  ylab('Accidents per 1000 People in 2019') +
  theme(legend.position = "none")
```

We can see that South Carolina has approximately double the accident rate of Oregon.

It's difficult to attribute this to some cause. It could be some problem with South Carolina's 
infrastructure or driving education but that lies outside the scope of this project.

We can see Iowa highlighted in red. Iowa is clearly not the worst in terms of accident rate, as 
many states have far worse rates, but there is also still room for improvement as there
are a few states with lower accident rates than Iowa.
<!-- End of Anders --->
  
  
  
<!-- Start of Jack --->

#Load Data Frames
```{r, message=FALSE}
census <- read_csv("/Users/jacknovak/Documents/DS202/ds202-group3-final/census.csv")%>%
  filter(STNAME == "Iowa", PRIMGEO_FLAG == 0, FUNCSTAT == "A")
iowa <- read_csv("/Users/jacknovak/Documents/DS202/ds202-group3-final/US_Accidents_Dec20.csv", col_types = cols(End_Lat=col_character(),End_Lng=col_character()))%>%
  select(State, Severity, Start_Time, Start_Lat, Start_Lng, `Wind_Speed(mph)`,Weather_Condition)%>%
  filter(State %in% "IA")

county <- map_data("county")
```

We also decided to look at accidents close to home. Evident in the map below, accidents in Iowa cluster in areas we know to be the most congested; namely, many accidents occur around Des Moines and along I-35 and I-80. Additionally, Iowa's northwestern corner appears to exhibit disproportionately many severe accidents relative to non-severe ones.

#Plot Accidents in Iowa
```{r}
iowacounties<- county%>%
  filter(region %in% "iowa")

p<-ggplot(iowacounties, aes(x=long, y=lat))+geom_polygon(aes(group=group))
p<-p+geom_point(data=iowa, aes(x=Start_Lng, y=Start_Lat, color=factor(Severity), alpha=0.15))
p<-p+labs(title="Accidents in Iowa", subtitle="Data from 2016 to 2020")+theme_map()
p
```

Nearly any Iowan can testify to the hazards of slippery weather. However, we sought to quantify weather's influence on accident frequency and severity by observing crashes recorded in Iowa over the past five years.

#Plot Severity by Wind Speed and Weather Type
```{r}
iowa1<-iowa%>%
  select(`Wind_Speed(mph)`,Severity, `Weather_Condition`)%>%
  na.omit()%>%
  group_by(Weather_Condition)%>%
  summarize(`Average Severity` = mean(Severity), `Average Wind Speed (mph)` = mean(`Wind_Speed(mph)`), n=n())%>%
  filter(n>15)
```

Slippery conditions are generally associated with the most severe crashes. Accidents in 'light freezing rain' were the most severe by a considerable margin, followed by snow, rain, and light thunderstorms. 

Based on the results, Iowans should limit driving on days with lots of precipitation and reduce their speeds. 

```{r}
ggplot(iowa1, aes(y=`Average Severity`, x=`Average Wind Speed (mph)`))+geom_text(aes(label=`Weather_Condition`), size=2)+labs(title="Severity and Wind Speed by Weather Type")
```

Iowans typically crash the most on weekdays. This result is pretty intuitive--people tend to drive more during the week.

#Accidents in Iowa by Day of the Week
```{r}
iowa %>%
  mutate(Day = wday(Start_Time, label=TRUE, abbr=FALSE))%>%
  group_by(Day)%>%
  ggplot(aes(x=Day))+geom_bar()+labs(title="Accidents in Iowa by Weekday")
```

<!-- End of Jack --->

Conclusion:
	
Although motor vehicle crashes are calamitous and will remain a problem into the future, our report reveals key signs of hope. Namely, we identified disparities in the number and severity of crashes across locations, suggesting states with very high crash rates may be able to reduce their crash rate through policy intervention.

Our data set supported the finding that the COVID-19 significantly increased the number and severity of motor vehicle crashes. We noted that the severity of crashes initially increased, then lulled, before increasing quite steeply at the end of 2020. This suggests that drivers drove faster and more recklessly after the start of the pandemic and that they did not adjust their behavior as cars returned to the roads.

Contribution Statement:

Anders:

Eric:

Jack: My analyses focused primarily on Iowa. I created a map of Iowa and two additional graphics to represent accidents by weekday and by weather type. I also contributed to this report as well as the project presentation.

Murat:

Ryan:


